Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-12 dev/paper/paper.tex
only findings left now

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting and focused
on single individuals: given some estimated model the goal is to find
valid counterfactuals for individual instance that fulfill various
desiderata. The ability of such counterfactuals to handle dynamics like
data and model drift remains a largely unexplored research challenge at
this point. There has also been surprisingly little work on the related
question of how the actual implementation of recourse by one individual
may affect other individuals. Through this work we aim to close that gap
by systematizing and extending existing knowledge. We first show that
many of the existing methodologies can be collectively described by a
generalized framework. We then argue that the existing framework fails
to account for a hidden external cost of recourse, that only reveals
itself when studying the endogenous dynamics of recourse at the group
level. Through simulation experiments involving various popular
counterfactual generators and several benchmark datasets, we generate a
total XX million Counterfactual Explanations and study the resulting
domain and model shifts. We find that the induced shifts are substantial
enough to likely impede the applicability Algorithmic Recourse in
practice. Fortunately, we find various potential mitigation strategies
that can be used in combination with existing approaches. Our simulation
framework for studying recourse dynamics is fast and open-sourced.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden, breakable, interior hidden]}{\end{tcolorbox}}\fi

@@

The dynamics illustrated in Figure~\ref{fig-poc} in
Section~\ref{sec-intro} were generated through a simple experiment that
aims to simulate the process of Algorithmic Recourse in practice. We
begin in the static setting at time time \(t=0\): given some pre-trained
classifier \(M\) we generate recourse for a random batch of \(B\)
individuals in the non-target class. Note that we focus our attention on
classification problems, since classification poses the most common
practical use-case for Algorithmic Recourse. In order to simulate the
dynamic process, we suppose that the model \(M\) is retrained following
the actual implementation of recourse in time \(t=0\). Following the
update to the model, we assume that at time \(t=1\) recourse is
generated for yet another random subset of individuals in the non-target
class. This process is repeated for a number of time periods \(T\). To
get a clean read on endogenous dynamics we keep the total population of
samples closed: we allow existing samples to move from factual to
counterfactual states, but do not allow any entirely new samples to
enter the population. The experimental setup is summarized in Algorithm
\ref{algo-experiment}

\begin{algorithm}
\caption{Simulation Experiment}\label{algo-experiment}
\begin{algorithmic}[1]
\Procedure{Experiment}{$M,D,G$}
\State $t\gets 0$
\While{$t<T$}
\State $D_B \subset D$
\State $D_B\gets G(D_B)$ \Comment{Generate counterfactuals.}
\State $M\gets M(D)$ \Comment{Retrain model.}
\EndWhile
\State \textbf{return} $M,D$
\EndProcedure
\end{algorithmic}
\end{algorithm}

A noteworthy practical consideration is the choice of \(T\) and \(B\).
The higher these values, the more factual instances undergo recourse
throughout the entire experiment. Of course, this is likely to lead to
more pronounced domain and model shifts by time \(T\). At the same time,
it is generally improbable that a very large part of the population
would request an explanation of the algorithm's decisions. In our
experiments, we choose the values such that \(T \cdot B\) corresponds to
the application of recourse on \(25-50\%\) of the negative instances
from the initial dataset. As we collect data at each time \(t\), we can
also verify the impact of recourse when it is implemented for a smaller
number of individuals.

Algorithm \ref{algo-experiment} summarizes the proposed simulation
experiment for a given dataset \(D\), model \(M\) and generator \(G\),
but naturally we are interested in comparing simulation outcomes for
different sources of data, models and generators. The framework we have
built facilitates this, making use of multi-threading in order to speed
up computations. Holding the initial model and dataset constant the
experiments are run for all generators, since our primary concern is to
benchmark different recourse methods. To ensure that each generator is
faced with the exact same initial conditions in each round \(t\), the
candidate batch of individuals from the non-target class is randomly
drawn from the intersection of all individuals in the non-target class
across all experiments
\(\left\{\textsc{Experiment}(M,D,G)\right\}_{j=1}^J\) where \(J\) is the
total number of generators.

@@

where \(\mathcal{F}\) is a unit ball in a Reproducing Kernel Hilbert
Space H {[}27{]}, and \(X\), \(\tilde{X}\) represent independent and
identically distributed samples drawn from probability distributions
\(p\) and \(q\) respectively {[}28{]}. MMD is a measure of the distance
between the kernel mean embeddings of \(p\) and \(q\) in RKHS
\(\mathcal{H}\). An important consideration is the choice of the kernel
function \(k(\cdot,\cdot)\). In our implementation we make use of the
radial basis function (RBF) kernel with a constant length-scale
parameter of \(0.5\). As RBF captures all moments of distributions \(p\)
and \(q\), we have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if
\(X=\tilde{X}\).

The evaluation metric in Equation~\ref{eq-mmd} is computed after every
round \(t=1,...,T\) of the experiment. To assess the statistical
significance of the observed shifts under the null hypothesis that
samples \(X\) and \(\tilde{X}\) were drawn from the same probability
distribution we follow
\protect\hyperlink{ref-arcones1992bootstrap}{{[}26{]}}. To that end, we
combine the two samples and generate a large number of permutations of
\(X + \tilde{X}\). Then, we split the permuted data into two new samples
\(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the
original samples. Then under the null hypothesis we should have that
\(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to
\(MMD_u^2[F,X,\tilde{X}]\). The corresponding \(p\)-value can then be
calculated by counting how these two quantities are not equal.

We calculate the MMD for both classes individually based on the ground
truth labels, i.e.~the labels that samples were assigned in time
\(t=0\). Throughout our experiments we generally do not expect the
distribution of the negative class to change over time -- application of
recourse reduces the size of this class, but since individuals are
sampled uniformly the distribution should remain unaffected. Conversely,
unless a recourse generator can perfectly replicate the original
probability distribution, we expect the MMD of the positive class to
increase. Thus, when discussing MMD, we generally mean the shift in the
distribution of the positive class.

Finally, \textbf{feature mean and feature standard deviation} are also
calculated to verify how the implementation of recourse impacts every
attribute in the dataset. Although MMD already captures information
about the expected value and variance, we may also be interested in a
more granular look at individual features.

@@

We extend the metric in Equation~\ref{eq-mmd} for the purpose of
quantifying model shifts. Specifically, we introduce \textbf{Predicted
Probability MMD (PP MMD)}: instead of applying Equation~\ref{eq-mmd} to
features directly, we apply it to the predicted probabilities assigned
to a set of samples by the model \(M\). If the model shifts, the
probabilities assigned to each sample will change; again, this metric
will equal 0 only if the two classifiers are the same. It is worth
noting that while we apply the technique to samples drawn uniformly from
the dataset, it can also be employed on arbitrary points in the entire
feature space (or a subspace). The latter approach is theoretically more
robust. Unfortunately, in practice this approachs suffers from the curse
of dimensionality, since it becomes increasingly difficult to select
enough points to overcome noise as the dimension \(D\) grows.

As an alternative to PP MMD we use a pseudo-distance for the
\textbf{Disagreement Coefficient} (Disagreement). This metric was
introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}27{]}} and
estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
classifiers do not agree on the predicted outcome for a randomly chosen
sample. Thus, it is not relevant whether the classification is correct
according to the ground truth, but only whether the sample lies on the
same side of the two respective decision boundaries. In our context,
this metric quantifies the overlap between the initial model (trained
before the application of recourse) and the updated model. A
Disagreement Coefficient unequal to zero is indicative of a model shift.
The opposite is not true: even if the Disagreement Coefficient is equal
to zero a model shift may still have occured. This is one reason for why
PP MMD is our our preferred metric.

@@

\hypertarget{sec-empirical-data}{%
\subsection{Data}\label{sec-empirical-data}}

We use three different real-world datasets from the Finance and
Economics domain, all of which are tabular and can be used for binary
classification. Firstly, we use the \textbf{Give Me Some Credit} dataset
which was open-sourced on Kaggle for the task to predict whether a
borrower is likely to experience financial difficulties in the next two
years \protect\hyperlink{ref-gmsc_data}{{[}28{]}}. Originally consisting
of 250,000 instances with 11 numerical attributes. Secondly, we use the
\textbf{UCI defaultCredit} dataset
\protect\hyperlink{ref-yeh2009comparisons}{{[}29{]}}, a benchmark
dataset that can be used to train binary classifiers to predict the
binary outcome variable, whether credit card clients default on their
payment. In its raw form it consists of 23 explanatory variables - 4
categorical features relating to demographic attributes\footnote{These
  have been ommitted from the analysis. See Section~\ref{sec-limit-data}
  for details.} and 19 continuous features largely relating to
individuals' payment histories and amount of credit outstanding. Both of
these datasets have been used in the literature on Algorithmic Recourse
before (see for example
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}} and
\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}), presumably
because they constitute real-world classification tasks involving
individuals that compete for access to credit.

As a third dataset we include the \textbf{California Housing} dataset
derived from the 1990 U.S. census and sourced through scikit-learn
\protect\hyperlink{ref-pace1997sparse}{{[}31{]}}. It consists of 8
continuous features that can be used to predict the median house price
for California districts. The continuous outcome variable is binarized
as \(\tilde{y}=\mathbb{I}_{y>\text{median}(Y)}\) indicating whether or
not the median house price of a given district is above or below the
median of all districts. While we have not seen this dataset used in the
previous literature on AR, others have used the Boston Housing dataset
in a similar fashion (see for example
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}). While we
initially also conducted experiments on that dataset, we eventually
discarded this dataset, since it has been found to suffer from an
ethical problem \protect\hyperlink{ref-carlisle2019racist}{{[}32{]}}.

\hypertarget{sec-empirical-classifiers}{%
\subsection{Classifiers and Generative
Models}\label{sec-empirical-classifiers}}

For each dataset and generator we look at three different types of
classifiers all of them built and trained using \texttt{Flux.jl}
\protect\hyperlink{ref-innes2018fashionable}{{[}33{]}}: firstly, a
simple linear classifier - \textbf{Logistic Regression} - implemented as
single linear layer with sigmoid activation; secondly, a multilayer
perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble}
composed of five MLPs following
\protect\hyperlink{ref-lakshminarayanan2016simple}{{[}34{]}} that serves
as our only probabilistic classifier. We have chosen to work with deep
ensembles both for their simplicity and effectiveness at modelling
predictive uncertainty. They are also the model of choice in
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. The actual neural
network architectures are kept simple (Table~\ref{tbl-mlp}), since we
are only marginally concerned with achieving good initial classifier
performance. For the real-world datasets we using mini-batch training
and dropout regularization.

@@

All classifiers and generative models are retrained for 10 epochs in
each round \(t\) of the experiment. Rather than retraining models from
scratch, we initialize all parameters at their previous levels (\(t-1\))
and compute backpropagate for 10 epochs using the new training data as
inputs into the existing model.

@@

@@

\hypertarget{potential-mitigation-strategies}{%
\subsection{Potential Mitigation
Strategies}\label{potential-mitigation-strategies}}

\hypertarget{sec-empirical-2-mitigate}{%
\subsubsection{Gravitational Counterfactual
Explanations}\label{sec-empirical-2-mitigate}}

A straight-forward choice simply extends the baseline approach by
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}: instead of
only penalizing the distance of the individuals' counterfactual to its
factual, we propose penalizing its distance to some sensible point in
the target domain, for example the sample average: \(\bar{\mathbf{x}}\).
For such a recourse objective, higher choices of \(\lambda_2\) relative
to \(\lambda_1\) will lead counterfactuals to gravitate towards the
specified point in the target domain. In the remainder of this paper we
will therefore refer to this approach as \textbf{Gravitational}
generator, when we investigate its potential usefulness for mitigating
endongenous macrodynamics\footnote{Note that despite the naming
  convention our goal here is not to provide yet another counterfactual
  generator, but merely investigate the most simple penalty we can think
  of with respect to its effectiveness.}.

\hypertarget{a-note-on-convergence}{%
\paragraph{A note on convergence}\label{a-note-on-convergence}}

For this simple mitigating strategy underlying the Gravitational
generator to work as expected, one needs to ensure that counterfactual
search continues, even after a predetermined threshold probability
\(\gamma\) has potentially already been reached.
Figure~\ref{fig-convergence} illustrates this distinction: if one
chooses to terminate search once the desired threshold is reached (left
panel) the gravitational pull towards \(\bar{\mathbf{x}}\) is never
actually satisfied (compare to right panel). More generally, if
convergence is defined simply in terms of flipping the predicted label
with some desired degree of confidence, this corresponds to essentially
ignoring any parts of the counterfactual search objective that do not
involve \(\ell(M(f(s_k^\prime)),t)\) beyond that point. While this may
be appropriate for some applications, in general this seems like an odd
convention. Since we nonetheless seen convergence specified simply in
terms of reaching the threshold probability in some places\footnote{\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}
  define convergence of Algorithm 1 in this way. The implementation of
  \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} in CARLA is
  also defined in this way.}, we thought it worth making this
distinction explicit.

@@

{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/gravitational_generator_comparison.png}

@@

\caption{\label{fig-convergence}Comparison of counterfactual search
outcome with simple (left) and strict convergence (right).}

@@

@@

For reasons stated earlier we have limited our analysis to
differentiable linear and non-linear classifiers, in particular logistic
regression and deep neural networks. While these sorts of classifiers
have also typically been analyzed in the existing literature on
Counterfactual Explanations and Algorithmic Recourse, they represent
only a subset of popular machine learning models employed in practice -
both black-box and glass-box. Despite the success and popularity of deep
learning in the context of high-dimensional data such as image, audio
and video, empirical evidence suggests that other models such as boosted
decision trees may have an edge when it comes to lower-dimensional
tabular datasets, such as the ones considered here
\protect\hyperlink{ref-grinsztajn2022tree}{{[}36{]}}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
\CSLLeftMargin{{[}26{]} }%
\CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
v statistics,''} \emph{The Annals of Statistics}, pp. 655--674, 1992.}

\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
\CSLLeftMargin{{[}27{]} }%
\CSLRightInline{S. Hanneke, {``A bound on the label complexity of
agnostic active learning,''} in \emph{Proceedings of the 24th
international conference on machine learning}, 2007, pp. 353--360.}

\leavevmode\vadjust pre{\hypertarget{ref-gmsc_data}{}}%
\CSLLeftMargin{{[}28{]} }%
\CSLRightInline{K. Competition, {``Give me some credit, improve on the
state of the art in credit scoring by predicting the probability that
somebody will experience financial distress in the next two years.''}
\url{https://www.kaggle.com/c/GiveMeSomeCredit}}

>>>>

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting and focused
on single individuals: given some estimated model the goal is to find
valid counterfactuals for individual instance that fulfill various
desiderata. The ability of such counterfactuals to handle dynamics like
data and model drift remains a largely unexplored research challenge at
this point. There has also been surprisingly little work on the related
question of how the actual implementation of recourse by one individual
may affect other individuals. Through this work we aim to close that gap
by systematizing and extending existing knowledge. We first show that
many of the existing methodologies can be collectively described by a
generalized framework. We then argue that the existing framework fails
to account for a hidden external cost of recourse, that only reveals
itself when studying the endogenous dynamics of recourse at the group
level. Through simulation experiments involving various popular
counterfactual generators and several benchmark datasets, we generate a
total XX million Counterfactual Explanations and study the resulting
domain and model shifts. We find that the induced shifts are substantial
enough to likely impede the applicability Algorithmic Recourse in
practice. Fortunately, we find various potential mitigation strategies
that can be used in combination with existing approaches. Our simulation
framework for studying recourse dynamics is fast and open-sourced.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, interior hidden, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi

@@

The dynamics illustrated in Figure~\ref{fig-poc} in
Section~\ref{sec-intro} were generated through a simple experiment that
aims to simulate the process of Algorithmic Recourse in practice. We
begin in the static setting at time time \(t=0\): given some classifier
\(M\) that was pre-trained on data
\(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\) we generate recourse
for a random batch of \(B\) individuals in the non-target class
(\(\mathcal{D}_0\)). Note that we focus our attention on classification
problems, since classification poses the most common practical use-case
for Algorithmic Recourse. In order to simulate the dynamic process, we
suppose that the model \(M\) is retrained following the actual
implementation of recourse in time \(t=0\). Following the update to the
model, we assume that at time \(t=1\) recourse is generated for yet
another random subset of individuals in the non-target class. This
process is repeated for a number of time periods \(T\). To get a clean
read on endogenous dynamics we keep the total population of samples
closed: we allow existing samples to move from factual to counterfactual
states, but do not allow any entirely new samples to enter the
population. The experimental setup is summarized in Algorithm
\ref{algo-experiment}

\begin{algorithm}
\caption{Simulation Experiment}\label{algo-experiment}
\begin{algorithmic}[1]
\Procedure{Experiment}{$M,\mathcal{D},G$}
\State $t\gets 0$
\While{$t<T$}
\State $\mathcal{D}_B \subset \mathcal{D}_0$ \Comment{Sample from $\mathcal{D}_0$.}
\State $\mathcal{D}_B\gets G(\mathcal{D}_B)$ \Comment{Generate counterfactuals.}
\State $M\gets M(\mathcal{D})$ \Comment{Retrain model.}
\EndWhile
\State \textbf{return} $M,\mathcal{D}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

A noteworthy practical consideration is the choice of \(T\) and \(B\).
The higher these values, the more factual instances undergo recourse
throughout the entire experiment. Of course, this is likely to lead to
more pronounced domain and model shifts by time \(T\). At the same time,
it is generally improbable that a very large part of the population
would request an explanation of the algorithm's decisions. In our
experiments, we choose the values such that \(T \cdot B\) corresponds to
the application of recourse on \(\approx50\%\) of the negative instances
from the initial dataset. As we collect data at each time \(t\), we can
also verify the impact of recourse when it is implemented for a smaller
number of individuals.

Algorithm \ref{algo-experiment} summarizes the proposed simulation
experiment for a given dataset \(\mathcal{D}\), model \(M\) and
generator \(G\), but naturally we are interested in comparing simulation
outcomes for different sources of data, models and generators. The
framework we have built facilitates this, making use of multi-threading
in order to speed up computations. Holding the initial model and dataset
constant the experiments are run for all generators, since our primary
concern is to benchmark different recourse methods. To ensure that each
generator is faced with the exact same initial conditions in each round
\(t\), the candidate batch of individuals from the non-target class is
randomly drawn from the intersection of all individuals in the
non-target class across all experiments
\(\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J\) where
\(J\) is the total number of generators.

@@

where \(\mathcal{F}\) is a unit ball in a Reproducing Kernel Hilbert
Space \(\mathcal{H}\)
\protect\hyperlink{ref-berlinet2011reproducing}{{[}26{]}}, and \(X\),
\(\tilde{X}\) represent independent and identically distributed samples
drawn from probability distributions \(\mathcal{X}\) and
\(\mathcal{\tilde{X}}\) respectively
\protect\hyperlink{ref-gretton2012kernel}{{[}27{]}}. MMD is a measure of
the distance between the kernel mean embeddings of \(\mathcal{X}\) and
\(\mathcal{\tilde{X}}\) in RKHS \(\mathcal{H}\). An important
consideration is the choice of the kernel function \(k(\cdot,\cdot)\).
In our implementation we make use of Gaussian kernel with a constant
length-scale parameter of \(0.5\). As the Gaussian kernel captures all
moments of distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\), we
have that \(MMD_u^2[F,X,\tilde{X}]=0\) if and only if \(X=\tilde{X}\).

The evaluation metric in Equation~\ref{eq-mmd} is computed after every
round \(t=1,...,T\) of the experiment. To assess the statistical
significance of the observed shifts under the null hypothesis that
samples \(X\) and \(\tilde{X}\) were drawn from the same probability
distribution, we follow
\protect\hyperlink{ref-arcones1992bootstrap}{{[}28{]}}. To that end, we
combine the two samples and generate a large number of permutations of
\(X + \tilde{X}\). Then, we split the permuted data into two new samples
\(X^\prime\) and \(\tilde{X}^\prime\) having the same size as the
original samples. Then under the null hypothesis we should have that
\(MMD_u^2[F,X^\prime,\tilde{X}^\prime]\) be approximately equal to
\(MMD_u^2[F,X,\tilde{X}]\). The corresponding \(p\)-value can then be
calculated by counting how these two quantities are not equal.

We calculate the MMD for both classes individually based on the ground
truth labels, i.e.~the labels that samples were assigned in time
\(t=0\). Throughout our experiments, we generally do not expect the
distribution of the negative class to change over time -- application of
recourse reduces the size of this class, but since individuals are
sampled uniformly the distribution should remain unaffected. Conversely,
unless a recourse generator can perfectly replicate the original
probability distribution, we expect the MMD of the positive class to
increase. Thus, when discussing MMD, we generally mean the shift in the
distribution of the positive class.

@@

We extend the metric in Equation~\ref{eq-mmd} for the purpose of
quantifying model shifts. Specifically, we introduce \textbf{Predicted
Probability MMD (PP MMD)}: instead of applying Equation~\ref{eq-mmd} to
features directly, we apply it to the predicted probabilities assigned
to a set of samples by the model \(M\). If the model shifts, the
probabilities assigned to each sample will change; again, this metric
will equal 0 only if the two classifiers are the same. We compute PP MMD
in two ways: firstly, we compute it over samples drawn uniformly from
the dataset, and, secondly, we compute it over points spanning a
meshgrid over a subspace of the entire feature space. For the latter
approach we bound the subspace by the extrema of each feature. While
this approach is theoretically more robust, unfortunately, it suffers
from the curse of dimensionality, since it becomes increasingly
difficult to select enough points to overcome noise as the dimension
\(D\) grows.

As an alternative to PP MMD we use a pseudo-distance for the
\textbf{Disagreement Coefficient} (Disagreement). This metric was
introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}29{]}} and
estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
classifiers do not agree on the predicted outcome for a randomly chosen
sample. Thus, it is not relevant whether the classification is correct
according to the ground truth, but only whether the sample lies on the
same side of the two respective decision boundaries. In our context,
this metric quantifies the overlap between the initial model (trained
before the application of recourse) and the updated model. A
Disagreement Coefficient unequal to zero is indicative of a model shift.
The opposite is not true: even if the Disagreement Coefficient is equal
to zero a model shift may still have occured. This is one reason for why
PP MMD is our our preferred metric.

@@

This section presents the exact ingredients and parameter choices
describing the simulation experiments we ran to produce the findings
presented in the next section (Section~\ref{sec-empirical-2}). For
convenience, we use Algorithm \ref{algo-experiment} as a template to
guide us through this section. A few high-level details upfront: each
experiment is run for a total of \(T=50\) rounds, where in each round we
provide recourse to five percent of all individuals in the non-target
class, so \(B_t=0.05 * N_t^{\mathcal{D}_0}\)\footnote{As mentioned in
  the previous section, we end up providing recourse to a total of
  \(\approx50\%\) by the end of round \(T=50\).}. All classifiers and
generative models are retrained for 10 epochs in each round \(t\) of the
experiment. Rather than retraining models from scratch, we initialize
all parameters at their previous levels (\(t-1\)) and compute
backpropagate for 10 epochs using the new training data as inputs into
the existing model. Evaluation metrics are computed and stored every 10
rounds.

\hypertarget{sec-empirical-classifiers}{%
\subsection{\texorpdfstring{\(M\) -- Classifiers and Generative
Models}{M -- Classifiers and Generative Models}}\label{sec-empirical-classifiers}}

For each dataset and generator we look at three different types of
classifiers all of them built and trained using \texttt{Flux.jl}
\protect\hyperlink{ref-innes2018fashionable}{{[}30{]}}: firstly, a
simple linear classifier - \textbf{Logistic Regression} - implemented as
single linear layer with sigmoid activation; secondly, a multilayer
perceptron (\textbf{MLP}); and finally, a \textbf{Deep Ensemble}
composed of five MLPs following
\protect\hyperlink{ref-lakshminarayanan2016simple}{{[}31{]}} that serves
as our only probabilistic classifier. We have chosen to work with deep
ensembles both for their simplicity and effectiveness at modelling
predictive uncertainty. They are also the model of choice in
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. The actual neural
network architectures are kept simple (Table~\ref{tbl-mlp}), since we
are only marginally concerned with achieving good initial classifier
performance. For the real-world datasets we using mini-batch training
and dropout regularization.

@@

@@

\hypertarget{sec-empirical-data}{%
\subsection{\texorpdfstring{\(\mathcal{D}\) --
Data}{\textbackslash mathcal\{D\} -- Data}}\label{sec-empirical-data}}

We use three different real-world datasets from the Finance and
Economics domain, all of which are tabular and can be used for binary
classification. Firstly, we use the \textbf{Give Me Some Credit} dataset
which was open-sourced on Kaggle for the task to predict whether a
borrower is likely to experience financial difficulties in the next two
years \protect\hyperlink{ref-gmsc_data}{{[}32{]}}. Originally consisting
of 250,000 instances with 11 numerical attributes. Secondly, we use the
\textbf{UCI defaultCredit} dataset
\protect\hyperlink{ref-yeh2009comparisons}{{[}33{]}}, a benchmark
dataset that can be used to train binary classifiers to predict the
binary outcome variable, whether credit card clients default on their
payment. In its raw form it consists of 23 explanatory variables - 4
categorical features relating to demographic attributes\footnote{These
  have been ommitted from the analysis. See Section~\ref{sec-limit-data}
  for details.} and 19 continuous features largely relating to
individuals' payment histories and amount of credit outstanding. Both of
these datasets have been used in the literature on Algorithmic Recourse
before (see for example
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}} and
\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}), presumably
because they constitute real-world classification tasks involving
individuals that compete for access to credit.

As a third dataset we include the \textbf{California Housing} dataset
derived from the 1990 U.S. census and sourced through scikit-learn
\protect\hyperlink{ref-pace1997sparse}{{[}35{]}}. It consists of 8
continuous features that can be used to predict the median house price
for California districts. The continuous outcome variable is binarized
as \(\tilde{y}=\mathbb{I}_{y>\text{median}(Y)}\) indicating whether or
not the median house price of a given district is above or below the
median of all districts. While we have not seen this dataset used in the
previous literature on AR, others have used the Boston Housing dataset
in a similar fashion (see for example
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}). While we
initially also conducted experiments on that dataset, we eventually
discarded this dataset, since it has been found to suffer from an
ethical problem \protect\hyperlink{ref-carlisle2019racist}{{[}36{]}}.

\hypertarget{g-generators}{%
\subsection{\texorpdfstring{\(G\) --
Generators}{G -- Generators}}\label{g-generators}}

All generators introduced earlier are included in the experiments:
Wachter \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
REVISE \protect\hyperlink{ref-joshi2019towards}{{[}8{]}}, CLUE
\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}, DiCE
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and Greedy
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. In addition, we
introduce two new generators in Section~\ref{sec-empirical-2-mitigate}
that directly address the issue of endogenous domain and model shifts.
We also test to what extent it may be beneficial to combine ideas
underlying the various generators.

@@

\hypertarget{sec-empirical-2-mitigate}{%
\subsection{Potential Mitigation
Strategies}\label{sec-empirical-2-mitigate}}

In the following we explore several ideas for strategies that may help
to mitigate those types of endogenous recourse dynamics we observed in
the previous subsection. All of them essentially boil down to one simple
principle: to avoid substantial domain and model shifts, the generated
counterfactuals should comply as much as possible with the true data
generating process. This principle is really at the core of Latent Space
generators, and hence it is not surprising that we have found these
types of generators to perform comparably well in the previous
subsection. But as we have mentioned earlier, generators that rely on
separate generative models carry an additional computational cost and -
perhaps more importantly - their performance hinges on the performance
of said generative models. Fortunately, it turns out that we can use a
number of other, much simpler strategies, which we will discuss now.

\hypertarget{more-conservative-decision-thresholds}{%
\subsubsection{More Conservative Decision
Thresholds}\label{more-conservative-decision-thresholds}}

The most obvious and trivial mitigation strategy is to simply choose a
higher decision threshold \(\gamma\). Under \(\gamma=0.5\),
counterfactuals will end up near the decision boundary by construction.
Since this is the region of maximal aleotoric uncertainty, the
classifier is bound to be thrown off. By simply setting a more
conservative decision threshold, we can avoid this issue to some extent.
A potential drawback of this approach is that a classifier with high
decisiveness may classify samples with high confidence even far away
from the training data.

\hypertarget{classifier-preserving-roar}{%
\subsubsection{Classifier Preserving
ROAR}\label{classifier-preserving-roar}}

Another potential strategy draws inspiration from ROAR
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}: to preserve the
classifier, we propose to simply explicitly penalize the loss it incurs
when evaluated on the counterfactual \(x^\prime\) at given parameter
values. Recall that \(g(\cdot)\) denotes what we had defined as the
external cost in Equation~\ref{eq-collective}. Then formally we let

\begin{equation}\protect\hypertarget{eq-clap}{}{
\begin{aligned}
g(f(s_k^\prime)) = l(M(f(s_k^\prime)),y^\prime)
\end{aligned}
}\label{eq-clap}\end{equation}

for each counterfactual \(k\) where \(l\) denotes the loss function used
to train \(M\). This approach, which we shall refer to as
\textbf{ClapROAR}, is based on the intuition that (endogenous) model
shifts will be triggered by counterfactuals that increase classifier
loss. It is closely linked to the idea of choosing higher decision
threshold, but likely better at avoiding the potential pitfalls
associated with highly decisive classifiers. It also makes the private
vs.~external cost trade-off more explicit and hence manageable.

\hypertarget{gravitational-counterfactual-explanations}{%
\subsubsection{Gravitational Counterfactual
Explanations}\label{gravitational-counterfactual-explanations}}

Yet another strategy simply extends Wachter as follows: instead of only
penalizing the distance of the individuals' counterfactual to its
factual, we propose penalizing its distance to some sensible point in
the target domain, for example the sample average: \(\bar{x}\):

\begin{equation}\protect\hypertarget{eq-clap}{}{
\begin{aligned}
g(f(s_k^\prime)) = \text{dist}(f(s_k^\prime),\bar{x})
\end{aligned}
}\label{eq-clap}\end{equation}

Once again we can putting this in the context of
Equation~\ref{eq-collective}, the former penalty can be thought of here
as the private cost incurred by the individual, while the latter
reflects the external cost incurred by other individuals. Higher choices
of \(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to
gravitate towards the specified point \(\bar{x}\) in the target domain.
In the remainder of this paper we will therefore refer to this approach
as \textbf{Gravitational} generator, when we investigate its potential
usefulness for mitigating endongenous macrodynamics\footnote{Note that
  despite the naming convention our goal here is not to provide yet
  another counterfactual generator, but merely investigate the most
  simple penalty we can think of with respect to its effectiveness.}.

Figure~\ref{fig-mitigation} shows an illustrative example that
demonstrates the differences in counterfactual outcomes when using the
various mitigation strategies compared to the baseline approach, that is
Wachter with \(\gamma=0.5\): choosing a higher decision threshold pushes
the counterfactual a little further into the target domain; this effect
is even stronger for ClapROAR; finally, using the Gravitational
generator the counterfactual ends up all the way inside the target
domain in the neighbourhood of \(\bar{x}\)\footnote{In order for the
  Graviational generator and ClapROAR to work as expected, one needs to
  ensure that counterfactual search continues, independent of the
  threshold probability \(\gamma\).}.

@@

{\centering \includegraphics[width=0.45\textwidth,height=\textheight]{www/mitigation.png}

@@

\caption{\label{fig-mitigation}The differences in counterfactual
outcomes when using the various mitigation strategies compared to the
baseline approach, that is Wachter with \(\gamma=0.5\)}

@@

Our findings \ldots{}

@@

For reasons stated earlier we have limited our analysis to
differentiable linear and non-linear classifiers, in particular logistic
regression and deep neural networks. While these sorts of classifiers
have also typically been analyzed in the existing literature on
Counterfactual Explanations and Algorithmic Recourse, they represent
only a subset of popular machine learning models employed in practice -
both black-box and glass-box. Despite the success and popularity of deep
learning in the context of high-dimensional data such as image, audio
and video, empirical evidence suggests that other models such as boosted
decision trees may have an edge when it comes to lower-dimensional
tabular datasets, such as the ones considered here
\protect\hyperlink{ref-grinsztajn2022tree}{{[}38{]}}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-berlinet2011reproducing}{}}%
\CSLLeftMargin{{[}26{]} }%
\CSLRightInline{A. Berlinet and C. Thomas-Agnan, \emph{Reproducing
kernel hilbert spaces in probability and statistics}. Springer Science
\& Business Media, 2011.}

\leavevmode\vadjust pre{\hypertarget{ref-gretton2012kernel}{}}%
\CSLLeftMargin{{[}27{]} }%
\CSLRightInline{A. Gretton, K. M. Borgwardt, M. J. Rasch, B. Schölkopf,
and A. Smola, {``A kernel two-sample test,''} \emph{The Journal of
Machine Learning Research}, vol. 13, no. 1, pp. 723--773, 2012.}

\leavevmode\vadjust pre{\hypertarget{ref-arcones1992bootstrap}{}}%
\CSLLeftMargin{{[}28{]} }%
\CSLRightInline{M. A. Arcones and E. Gine, {``On the bootstrap of u and
v statistics,''} \emph{The Annals of Statistics}, pp. 655--674, 1992.}
